---
title: "Data Visualization - Project 1"
author: "Stefano Meschiari"
date: "August 1, 2015"
output: 
  html_document: 
    fig_height: 6
    fig_width: 8
---
This document contains two visualizations of the data contained in [this dataset from the GISTEMP site](ExcelFormattedGISTEMPData2CSV.csv). The original dataset is described [here](http://data.giss.nasa.gov/gistemp/). 

Briefly, the original dataset contains temperature anomalies (averaged over the year, measured in units of 0.01 degrees Celsius) as a function of time (year) and latitude. Temperature measurements are scaled against the 1950-1981 period. 

This document contains both the visualizations, and the R code used to produce them. Each plot has an interactive version hosted on plot.ly.

## Yearly global temperature anomaly 
#### [Interactive Version](https://plot.ly/~exostefanom/31.embed)
<center>
```{r, echo=FALSE, cache=TRUE, message=FALSE, warnings=FALSE}
library(dplyr)
library(ggplot2)
library(plotly)
library(readr)
library(tidyr)
library(latex2exp)
data <- read_csv('data.csv') %>%
  gather(Position, TemperatureAnomaly, -Year) %>%
  mutate(TemperatureAnomaly = TemperatureAnomaly / 100)

ggplot(data=filter(data, Position=="Glob")) +
  geom_line(aes(Year, TemperatureAnomaly), color='gray') +
  geom_point(aes(Year, TemperatureAnomaly, fill='Data')) +
  geom_smooth(aes(Year, TemperatureAnomaly, color="Fit")) +
  ylab("Global temperature anomaly (°C)") +
  ggtitle("Yearly temperature anomaly (global)") +
  guides(color = guide_legend(element_blank()),
               fill = guide_legend(element_blank())) +
  ylab(latex2exp("Global temperature anomaly ($\\degree$C)")) +
  theme(legend.position=c(0.9, 1.05), legend.box="horizontal", 
              plot.title=element_text(hjust=0, vjust=1))
```
</center>
The plot above shows the global world temperature anomaly (in degrees Celsius) on the y-axis and time (years) on the x-axis. 

Each data point is represented as a black dot. Overplotted is a smoothed version of the same data (red line), which shows the overall behavior of the temperature anomaly over time by smoothing over the small inter-year variations. The smoothing was done by fitting a LOESS function over groups of 80 year measurements. The gray shade represents the confidence intervals of the fit.

This plot shows a prominent increase in global temperature as a function of time, with an accelerating temperature growth post-1960s.

## Yearly global temperature anomaly, by latitude
#### [Interactive Version](https://plot.ly/~exostefanom/50.embed)
<center>
```{r, echo=FALSE, cache=TRUE, message=FALSE, warnings=FALSE}

data2 <- read_csv('data.csv') %>%
  select(-(Glob:`90S-24S`)) %>%
  gather(Position, TemperatureAnomaly, -Year) %>%
  mutate(TemperatureAnomaly = TemperatureAnomaly / 100)

ggplot(data=data2) +
  geom_tile(aes(x=Year, y=Position, fill=TemperatureAnomaly)) +
  scale_fill_distiller(type='div', palette="RdBu", trans="reverse",
                       guide=guide_colorbar(title=element_text("Temperature\nanomaly (°C)"), direction="horizontal")) +
  scale_y_discrete(limits=rev(levels(data2$Position))) +
  ylab("Latitude") +
  ggtitle("Yearly temperature anomaly (by latitude)") +
  theme(legend.position=c(0.8, 1.05), legend.box="horizontal",
              plot.title=element_text(hjust=0, vjust=1))
```
</center>
The plot above shows the temperature anomaly (in degrees Celsius) as a function of time (years) and latitude (°). 

The temperature anomaly is plotted as a heat map, spanning from -2°C to +2°C. Latitude (on the y-axis) is divided into 8 chunks spanning from 90° south to 90° north.

This plot shows that the inter-year temperature variations are stronger at the highest latitudes (64°-90°), with the highest temperature anomaly peaks reached in the North Pole (top band). 